Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
changes on abstract. Left some comments for Clement

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%% Author with two affiliations and emails.
\author{Cl\'ement B\'era}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{clement.bera@vub.ac.be}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Elisa Gonzalez Boix}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{egonzale@vub.ac.be}          %% \email is recommended

@@

\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".  The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

As the VM has evolved, including adding a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

%% Keywords
%% comma separated list
\keywords{Just-in-Time compiler, Virtual machine, Managed runtime, Tools}  %% \keywords are mandatory in final camera-ready submission

>>>>

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%% Author with two affiliations and emails.
\author{Cl\'ement B\'era}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{clement.bera@vub.be}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Elisa Gonzalez Boix}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{egonzale@vub.be}          %% \email is recommended

@@

\egb{The title is a bit misleading. It sounds great for a journal paper on the \openSmalltalkVM, but probably it is best to adapt it or add a subtitle on the specific topic you touch upon in this paper}

\egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution. I left the original text in comments below}

\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. While simulation is used to develop and debug the VM, production is used to release the VM.

%\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

\egb{I think you can skip the next paragraph from the abstract. If the point is that you need to co-evolve both execution models once the VM changes, you can do that with 1 line in the previous paragraph. Now it may look like you will discuss co-evolution issues, but I guess you do not}

As the VM has evolved, including the addition of a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

%TODO egb: revisit this paragraph after reading all paper.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

%% Keywords
%% comma separated list
\keywords{Just-in-Time compiler, virtual machine, managed runtime, tools}  %% \keywords are mandatory in final camera-ready submission
